---
title: "Activity_Monitoring_markdown"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Loading the raw data file into R and installing the required packages

```{r Chunk1}
activity_monitoring_data <- read.csv("activity.csv", header=TRUE)
install.packages("dplyr", repos = "http://cran.us.r-project.org", dependencies = TRUE)
install.packages("timeDate", repos = "http://cran.us.r-project.org", dependencies = TRUE)
install.packages("reshape2", repos = "http://cran.us.r-project.org", dependencies = TRUE)
install.packages("ggplot2", repos = "http://cran.us.r-project.org", dependencies = TRUE)
library(timeDate)
library(reshape2)
library(ggplot2)
library(dplyr)
```

#Plotting the Histogram of the total number of steps taken each day

```{r Chunk2}
#Calculate total steps per day and convert into a dataframe
total_steps <-with(activity_monitoring_data, tapply(steps,date,sum, na.rm=TRUE))
total_steps_df <-melt(total_steps, varnames = "date", value.name = "total_steps")
rm(total_steps)
with(total_steps_df, barplot(total_steps, names.arg=date, xlab="Date", ylab="Total Steps", main="Histogram of Total Steps taken each day in the month of October and November"))
```

#Calculating mean and median number of steps taken each day

```{r Chunk3}
summary_data <- activity_monitoring_data %>% group_by(date) %>% summarise(mean_steps=mean(steps,na.rm=TRUE),
                                                                          median_steps=median(steps[steps>0], na.rm=TRUE))

par(mfrow=c(1,2))

#Plotting the Histogram of Mean and Median steps for each day in October and November

with(summary_data, barplot(mean_steps, names.arg=date, xlab="Date", ylab="Mean Steps", main="Histogram of Mean steps each day"))

with(summary_data, barplot(median_steps, names.arg=date, xlab="Date", ylab="Median Steps", main="Histogram of Median steps each day"))
```

#Plotting a time series plot using the Summary data from Chunk3 for mean steps
```{r Chunk4}
ggplot(summary_data, aes(as.Date(date), mean_steps)) + geom_line(na.rm=TRUE)
```

#Calculate the 5 minute interval with maximum average number of steps
```{r Chunk5}
summary_data <- activity_monitoring_data %>% group_by(interval) %>% summarise(mean_steps=mean(steps,na.rm=TRUE)) %>% arrange(-mean_steps) %>% slice(n=1)
interval<- summary_data$interval
mean_steps <- summary_data$mean_steps
## Interval 835 has maximum mean steps = 206
```

# Imputation for missing values
```{r Chunk6}
#Replace the missing value for steps column by average value for each interval to include the impact of each interval on the
#number of steps taken

activity_monitoring_data <- activity_monitoring_data %>% group_by(interval) %>% mutate(steps_impute =ifelse(is.na(steps),mean(steps, na.rm=TRUE), steps))
```

# Histogram of total steps each day using new imputed steps column in Chunk6
```{r Chunk7}
total_steps <-with(activity_monitoring_data, tapply(steps_impute,date,sum, na.rm=TRUE))
total_steps_df <-melt(total_steps, varnames = "date", value.name = "total_steps")
rm(total_steps)
with(total_steps_df, barplot(total_steps, names.arg=date, xlab="Date", ylab="Total Steps", main="Histogram of Total Steps taken each day after imputation in the month of October and November"))
```

# Panel plot for steps in each 5minute interval between weekdays and weekends
```{r Chunk8}
activity_monitoring_data$day <- as.factor(ifelse(isWeekday(as.Date(activity_monitoring_data$date)), "weekday", "weekend"))
summary_data <- activity_monitoring_data %>% group_by(interval, day) %>% summarise(mean_steps=mean(steps,na.rm=TRUE))
ggplot(summary_data, aes(interval,mean_steps)) + geom_bar(stat='identity') + labs(x="5 Minute Interval", y="Mean Steps", title="Histogram of Mean Steps taken in each 5 Minute Interval compared between Weekdays & Weekends") + facet_wrap(~day)
```

